Allowing Stable Groups of Prosumers to Enter the Market through Proper Aggregation

### 2014-08-22 Article.tex
update

\begin{abstract}
%\boldmath
There is a growing interest for coalitional concepts in the smart grid, spanning multiple objectives such as power losses reduction, benefits maximization or market stability... In this paper, we study coalitions of prosumers (agents that produce but also consume depending on time) which aim at selling energy to the grid. We reckon that the grid is able to specify two requirements (stability and minimum production) that enable a coalition to enter the market. Based on probability distributions of units available power, coalitions which are able to achieve low standard deviations are considered as stable coalitions. As it seems intuitively understandable that maintaining stable coalitions would be less expensive in terms of communication load than highly versatible ones, our objective consists in forming coalitions that satisfy grid requirements both in terms of stability and minimum production. Our definition of stability being implicitely linked to correlation between agents timeseries (see section 3), we organize them over a correlation graph and use a variation of clique percolation to construct valid coalitions in reasonable time.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

\section{Introduction}
% no \IEEEPARstart
One of the most basic ideas in the smart grid revolves around the introduction of communication means in the power grids, which could enable complex improvements in the energy management and lead progressively to a greener energetic system [1] [2]. Distributed energy resources (DER) such as wind turbines or photovoltaic panels are supposed to emerge and populate not only spread out farms, but also classical residential neighborhoods. Together with electric vehicles, and demand side management tools, they constitute the building blocks which will help turn the today pure energy consumers into true actors of the grid operation [3]. Such energy aware agents that consume, produce, and are ready to make concessions (appliances delays, V2G…) to ensure grid stability are commonly called prosumers \cite{Rathnayaka2012}.

@@

\begin{itemize}
\item Realistic representations (prosumers productions are non independent variables and come (indirectly) from real traces)
\item Minimizing the standard deviations of coalitions production probability distributions through the use of a proper utility function (see section 3)
\item Being (relatively) scalable with the number of prosumers
\end{itemize}

In more details, and as will be explained in section 3, we consider agents that, depending on meteorological conditions, personal preferences, and their appliances (loads and generators), consume and produce more or less energy (using only wind turbines and PV as generators, which can be quite simply and efficiently modeled).  We used real meteorological traces [13] to account for seasonal and daily variations in the prosumers energetic output profiles as well as realistic correlations. With these ingredients, we are able to run simulations and record the different output profiles as timeseries.

As correlations play a significant role in our mechanism, we organize the prosumers in a correlation graph, a quite popular approach in market stocks clustering [16] [17] (see section 2). We will see that in such graphs, cliques represent structures of interest. In our settings, it means that cliques are “good” (from the utility point of view) foundations for high valued coalitions. We thereby use a modified clique percolation algorithm (see section 4) that will enable us to expand the cliques as much as needed in order to form proper coalitions that fulfill grid requirements.

The rest of the paper is organized as follows, section 2 locates our work in the related litterature. Section 3 explains the model we used to obtain realistic prosumers available power timeseries and introduces some notations. Section 4 relates the method for building coalitions based on the agent timeseries. Finally, section 5 provides some results.

\section{Related Work}

@@

In this paper and as will be explained in section 3, coalitions utilities depend on statistical properties of historical values. The goal being to form coalitions that, statistically spealing, are more likely to exhibit stable behaviors. We will see that, when it comes to the formation of the coalitions, a prosumer is completely interchangeable with the timeserie representing its available production.

The setting is thus similar with some fiancial studies on stock exchanges, where researchers tried to find relevant clusters of stocks based on their daily prices variations. The problem of clustering is usually approached by means of a similarity measure and completed by a clustering technique such as K-means or hierarchical clustering. Nevertheless, in \cite{Mantegna1999} \cite{Garas2008}, stocks are represented by the timeseries of their daily log returns. Quite intuitive similarity measures based on Pearson correlation coefficient are established ($ d_{ij} = \dfrac{1}{2}\sqrt{2(1-\rho_{ij})}\ or\ d_{ij} = 1 - \rho_{ij}^{2} $ ) such that correleted stocks are close to each other. Such a situation can be represented by a complete graph $ G = (V,E,\omega) $ such that a vertex stands for a stock and the weight of the edge connecting two stocks i and j is equal to $ d_{ij} $. However, in this complete state, the graph G is of little use as precious information is flooded. Several filtering techniques have been proposed, among which, minimum spanning tree \cite{Mantegna1999}, k-nearest graphs [], or $\epsilon$-graphs [][].

The idea of $\epsilon$-graphs consists in filtering edges based on their weight, only keeping edges whose weights are less than $ \epsilon $. Despite the procedure simplicity, the choice of $ \epsilon $ (or k for k-nearest graphs) is not trivial and may influence strongly the results. There is indeed to our knowledge no optimal rule for chosing $ \epsilon $. In \cite{Garas2008}, the authors studied the topological properties (mainly average clustering and connectivity) of the correlation graph depending on the threshold $ \epsilon $. One of the conclusions was that "strong links" (i.e links between strongly correlated timeseries) are responsible for clustering while "weak links" provide network connectivity (in such a network, weak links are indeed connecting densely connected communities).

Presented in this way, the timeseries clustering task is also very close to graph community detection. Communities in networks are indeed often seen as groups of nodes exhibiting high internal densities of links as well as a low density across communities \cite{Newmanb}. Although several techniques exists (based on different graph properties : modularity \cite{Newmanb}, edge betweeness \cite{Girvan2002}, spectrum of Laplacian \cite{Newman}...), the clique percolation algorithm \cite{Lancichinetti} uses directly this observation and the idea behind it is actually quite simple. It starts indeed by searching for cliques (a complete subgraph) in the given graph as potential community cores. Because other nodes can naturally belong to a community with a less strong connectivity requirement, the algorithm uses a fitness function that quantifies the quality (according to some criterions) of a community structure inside the graph. Expansed cliques alternatively look in their respective neighborhood, select the node that yields the best increase in fitness, and finally incorporate it. The percolation algorithm stops when seeds cease to grow or when all nodes are affected to at least one seed. At this point, some coalitions may be partially overlapping, meaning that a node can belong to multiple communities. Detection of overlapping communities is actually a very active field of research, especially in social networks where persons often belongs to several communities (family, friends, work colleagues...).

@@

\section{Model}
A major concern while designing our model was its ability to translate correctly realistic patterns (consumption and production) as well as realistic geographical correlations between patterns. Largely because of these reasons (but also because they are easily, and sometimes freely, available online) we chose to use meteorological traces as inputs. In this paper we used namely both french data from 2006 to 2012 sampled every 3 hours \cite{Infoclimat} (similar data for the United States covering year 2010 can also be found at \cite{NCDC}). The weather stations already offer a quite good sampling of a given territory, and, with the growth of personal weather stations constantly updating data bases, the mesh becomes finer and finer. As shown in the first part of figure 1, the first step consists in discretizing the studied zone around well chosen weather stations and gathering traces for these stations. In this paper, we consider that prosumers can only produce through wind turbines and photovoltaic panels (PV). On the other hand, we built their consumption patterns according to two major cycles :
\begin{itemize}
\item \textbf{Daily cycles} : Consumption is low during night, and higher during the day with two picks in the morning and evening. Some noise is added so that prosumers have similar but not identical cycles.
\item \textbf{Seasonal cycles} : Consumption is higher in the winter because of heating and low in the summer (we did not consider air conditioning in this paper). Temperature traces were the principal ingredient for modeling these cycles.
\end{itemize}

\begin{center}
\begin{figure}
\includegraphics[scale=0.45]{figure2.jpg}
\caption{Process diagram}
\end{figure}
\end{center}

@@

At this point, modeling agents consists in fixing a few parameters (most of the time drawn from random distributions) such as the geographical position, the number of wind turbines, PV, appliances, the temperature of confort, and so on... The objective is that, depending on the weather of her zone, the DER and appliances she owns, and the way she decides to heat her home, a given prosumer is able to compute her production/consumption at any time. We thus used simple and known power curve models for linking weather variables to output power, such that a wind turbine (or a PV for instance) takes a wind speed (a solar irradiance) as input and gives an instantaneous output power [] []. Any agent is thus able to compute at any time (within the traces bounds) how much she produces and consumes.

@@

More formally, we denote by $ \mathcal{A} = \{ a_{1},...a_{N} \} $ the set of agents (the prosumers) and $ P_{i}(t) $ the instantaneous power value of agent i at instant t (its instantaneous production minus consumption, meaning that $ P_{i}(t) $ represents agent i available power at instant t). During the simulation (from $t_{0} $ to $ t_{K} $), all agents record their values with an hour time interval. As expected (because we introduced them), the timeseries exhibit high seasonal patterns, but completely different from agent to agent because they depend both on the energetic mix and habits of a given prosumer. Nevertheless, these macro variations can hide the interesting information in the correlation coefficients and twist the rest of our process. Thereby, we remove these seasonal effects for all agents (see second block of figure 1). We denote by $ \mathcal{T}_{i} = \{ P_{i}(t_{0}),...,P_{i}(t_{K}) \} $ the resulting timeserie for  agent i. For any agent i, we note $ \mathcal{P}_{i} $ the probability distribution drawn from $ \mathcal{T}_{i} $ and by $ P_{i} $ a random variable following $ \mathcal{P}_{i} $.

@@

What this paper investigates in the following is the developement of an algorithm that organizes prosumers such that the synergy term of the standard deviation ( $ \rho_{ij}\sigma_{i}\sigma_{j} $ in the example above) is minimized. In such settings, and through the clique percolation procedure (section 4), we will show that coalitions with low relative standard deviation, e.g high utility coalitions, can be computed.

\section{Forming coalitions}

This section explains the process with which we form the coalitions (see the third block of Fig. 1). First, we need to simulate the timeseries of available power (first two blocks of Fig. 1). We consider a pool $ \mathcal{A} $ of 200 agents, whose parameters were chosen randomly. The prosumers are positioned (also randomly) on a square lattice previously filled with climate vectors obtained from the french data sets (see section 3). Simulations were run from february 2006 to december 2010 such that we are dealing with 200 hourly sampled timeseries of available power over this period. Removing season trends finally leads us to the formation of coalitions.

The model we used in order to simulate timeseries of available production provide some diversity because of the energetic mix/climate vectors combo. Nevertheless, as the number of agents grows, the timeseries tend to exhibit similar patterns. This is particularly visible when creating a correlation graph $ G_{1}(\mathcal{A},E_{1},\omega_{1}) $ with the same kind of metric as [] or [] ($ d_{ij} = 1 - \rho_{ij}^{2} $). Very well defined clusters appear in the corresponding $ \epsilon $-graph for large range of $ \epsilon $.

@@

We thus opt for reversing the metric ($ d_{ij}^{'} = \rho_{ij}^{2} $) such that uncorrelated timeseries are close to each other and (anti)correlated timeseries are distant in $ G_{2}(\mathcal{A},E_{2},\omega_{2}) $. As expected [], independently of the $ \epsilon $ parameter, the resulting $ \epsilon $-graph exhibits henceforth much less clustering than in $ G_{1} $ and communities seem hardly visible. Therefore, using classical clustering or community detection algorithms do not provide very good results.

@@

As explained in section 2, clique percolation leads generally to overlapping communities. For simplicity, in this paper, we wish to keep the coalitions separated and  leave the management of overlapping coalitions for future work. We thus implemented a simple heuristic that consider nodes in multiple seeds one by one and chooses its final coalition as the one that “needs it the most” in terms of utility loss. More formally, for a coalition S and a node $ i \in S $, we define :

@@

\section{Results}

\begin{center}
\begin{figure}
\begin{tabular}{cc}
   \includegraphics[scale=0.44]{./figure1/util.eps} &
   \includegraphics[scale=0.44]{./figure3/utility.eps} \\
\end{tabular}
\caption{Utility function in parameters space}
\end{figure}
\end{center}

The first two parameters ($\phi$ and $P_{\phi}^{MIN}$) shape the utility function such that, if $ \phi $ is close to zero, the reliability requirement is very high and only small values of $ P_{\phi}^{MIN}$ could lead to valid coalitions (and positive utility values). Conversely, the higher $\phi$, the less constraints are imposed to the coalitions and valid ones can arise for a larger spectrum of $ P_{\phi}^{MIN}$ (see figure 1, left plot).

@@

As is visible on the right plot of figure 1, when grid constraints are fixed and only the number of coalitions varies, we can see an increase in social welfare (sum of coalitions utilities) up to a maximum point before a decrease. The reason is that increasing the number of coalitions allows more coalitions to reach stability and enter the market, but there is a point where nodes bringing stability are not sufficient inside the coalitions to make them pass the grid requirements, and some coalitions start to fail with zero utility. Moreover, reckon that increasing $ N_{COAL} $ means also increasing $ \epsilon $, leading to graphs where information is flooded, meaning that the algorithm performances decrease.
Naturally, it is also visible that small values of $P_{\phi}^{MIN}$ lead to higher utilities because coalitions are able to announce higher contract values.

\begin{center}
\begin{figure}
  \includegraphics[scale=0.8]{./figure5/coals.pdf}
  \caption{Evolution of the coalitions for different values of grid requirement $ P_{MIN} $. Blue dots represents clique percolation results while red dots stands for random process results.}
\end{figure}
\end{center}

For our 200 prosumers, we suppose that the grid fixes a reliability $ \phi = 0.1 $ (meaning that coalitions should produce more than their contract values at least $ 90 \% $ of the time. We also fix the number of coalitions to 10 and have a look at how these coalitions evolve as the grid changes the $ P_{MIN} $ requirement. As a comparison, we use a completely random algorithm that only asks for a number of coalitions and partitions the agents in a random fashion. Figure 3 shows this evolution for our algorithm (blue dots) and for the random process (red dots). The diameter of a dot is proportional to the number of agents in the coalition and the higher the dot is, the higher its utility. The $ P_{MIN} $ values of the x axis are expressed in tenth of MW for readability and the hatched zone corresponds to the "under-requirement space", meaning that whenever a coalition is in this zone, it has a null utility.

Looking at figure 3, we can see first that our algorithm seems to make generally a better job at finding high valued coalitions, but also that the results seem far more robust against how the grid positions its requirements. The blue dots allowed to enter the market seem indeed to outnumber the red dots, especially when the grid requirements are neither to low or to high ($ P_{MIN} = 5 $ in figure 3).

In order to explore a little bit more in this direction, we have a look, on figure 4, at the percentage of coalitions able to enter the market for different values of $ P_{MIN} $. For consistency, we always average the results of the random procedure over 100 realizations (5 realizations for our algorithm, but the results are relatively stable), and the errorbars in the plots stands for the standard deviations of the results.

@@

\begin{center}
\begin{figure}
   \includegraphics[scale=0.7]{./figure4/survival.eps}
   \caption{Percentage of coalitions able to enter the market for different values of grid requirement $ P_{MIN} $. Blue curve represents clique percolation results while red curve stands for random process results}
\end{figure}
\end{center}

@@

\section{Conclusion}
In this paper we collected meteorological traces for weather stations sampling a given territory. After discetization of the space, each zone receives the climate vector corresponding to its mother weather station. Prosumers are then distributed randomly between zones and each prosumer receives a random number of DER and loads as well as diverse consumption habits. Simulations are then runned and allow us to obtain, for each prosumer, the timeserie describing the instantaneous power that he could potentially be wanted to sell on the market. Nevertheless, as entering on the energy market is constrained by the grid operator, such that only sufficiently reliable and sufficiently productive units are allowed to enter, prosumers have usually no other choice than forming coalitions. We then introduced a framework such that the grid requirements take a precise meaning and such that coalitions announce contract values accordingly.

Through this paper, our interest was on how we could circumscribe groups of prosumers that can rise above the grid requirements. In this direction, we removed seasonal variations from the prosumers timeseries and proposed to organize them over a correlation graph with a reversed metric. A greedy clique percolation process on the $ \epsilon $-graph enables us to finally establish coalitions of agents. We showed in the last section that our algorithm was actually finding coalitions with high utility values but also that it was far more robust than a random search over coalition structures when the grid operator increases his requirements (Fig. 4).

>>>>

\begin{abstract}
%\boldmath
There is a growing interest for coalitional concepts in the smart grid, spanning multiple objectives such as power losses reduction, benefits maximization or market stability... In this paper, we study coalitions of prosumers (agents that produce but also consume depending on time) which aim at selling energy to the grid. We reckon that the grid is able to specify two requirements (stability and minimum production) that enable a coalition to enter the market. Based on probability distributions of units available power, coalitions which are able to achieve low standard deviations are considered as stable coalitions. As it seems intuitively understandable that maintaining stable coalitions would be less expensive in terms of communication load than highly versatible ones, our objective consists in forming coalitions that satisfy grid requirements both in terms of stability and minimum production. Our definition of stability being implicitely linked to correlation between agents timeseries (see section \ref{sec:model}), we organize them over a correlation graph and use a variation of clique percolation to construct valid coalitions in reasonable time.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

\section{Introduction}
\label{sec:introduction}
% no \IEEEPARstart
One of the most basic ideas in the smart grid revolves around the introduction of communication means in the power grids, which could enable complex improvements in the energy management and lead progressively to a greener energetic system [1] [2]. Distributed energy resources (DER) such as wind turbines or photovoltaic panels are supposed to emerge and populate not only spread out farms, but also classical residential neighborhoods. Together with electric vehicles, and demand side management tools, they constitute the building blocks which will help turn the today pure energy consumers into true actors of the grid operation [3]. Such energy aware agents that consume, produce, and are ready to make concessions (appliances delays, V2G…) to ensure grid stability are commonly called prosumers \cite{Rathnayaka2012}.

@@

\begin{itemize}
\item Realistic representations (prosumers productions are non independent variables and come (indirectly) from real traces)
\item Minimizing the standard deviations of coalitions production probability distributions through the use of a proper utility function (see section \ref{sec:model})
\item Being (relatively) scalable with the number of prosumers
\end{itemize}

In more details, and as will be explained in section \ref{sec:model}, we consider agents that, depending on meteorological conditions, personal preferences, and their appliances (loads and generators), consume and produce more or less energy (using only wind turbines and PV as generators, which can be quite simply and efficiently modeled).  We used real meteorological traces [13] to account for seasonal and daily variations in the prosumers energetic output profiles as well as realistic correlations. With these ingredients, we are able to run simulations and record the different output profiles as timeseries.

As correlations play a significant role in our mechanism, we organize the prosumers in a correlation graph, a quite popular approach in market stocks clustering [16] [17] (see section \ref{sec:related}). We will see that in such graphs, cliques represent structures of interest. In our settings, it means that cliques are “good” (from the utility point of view) foundations for high valued coalitions. We thereby use a modified clique percolation algorithm (see section \ref{sec:results}) that will enable us to expand the cliques as much as needed in order to form proper coalitions that fulfill grid requirements.

The rest of the paper is organized as follows, section \ref{sec:related} locates our work in the related litterature. Section \ref{sec:model} explains the model we used to obtain realistic prosumers available power timeseries and introduces some notations. Section \ref{sec:forming} relates the method for building coalitions based on the agent timeseries. Finally, section \ref{sec:results} provides some results.

\section{Related Work}
\label{sec:related}

@@

In this paper and as will be explained in section \ref{sec:model}, coalitions utilities depend on statistical properties of historical values. The goal being to form coalitions that, statistically spealing, are more likely to exhibit stable behaviors. We will see that, when it comes to the formation of the coalitions, a prosumer is completely interchangeable with the timeserie representing its available production.

The setting is thus similar with some fiancial studies on stock exchanges, where researchers tried to find relevant clusters of stocks based on their daily prices variations. The problem of clustering is usually approached by means of a similarity measure and completed by a clustering technique such as K-means or hierarchical clustering. Nevertheless, in \cite{Mantegna1999} Mantegna introduced another approach where stocks, represented by the timeseries of their daily log returns, are organized in a graph such that stocks (vertices) exhibiting similar price fluctuation patterns are close to each other. This closeness notion is formalized with a similarity measure based on Pearson correlation coefficient ($ d_{ij} = \dfrac{1}{2}\sqrt{2(1-\rho_{ij})} $ but sometimes also $ d_{ij} = 1 - \rho_{ij}^{2} $ ) that enables to weight the edges of the graph. Because there is a distance between every two nodes, \cite{Mantegna1999} obtains a complete graph where precious information is flooded. In order to discard unsignificant information, \cite{Mantegna1999} used a minimum spanning tree oriented technique and achieves a hierarchical clustering of the stocks. Since then, several other filtering techniques have been proposed, such as k-nearest graphs, or $ \epsilon $-graphs, that relax the tree assumption \cite{Onnela2004}.

The idea of $\epsilon$-graphs consists in filtering edges based on their weight, only keeping edges whose weights are less than $ \epsilon $. Despite the procedure simplicity, the choice of $ \epsilon $ (or k for k-nearest graphs) is not trivial and may influence strongly the results. There is indeed to our knowledge no optimal rule for chosing $ \epsilon $. In \cite{Garas2008} and \cite{Onnela2004}, the authors studied the topological properties (average clustering, connectivity, relative number of cliques...) of the correlation graph against those of growing random graphs, depending on the threshold $ \epsilon $. One of the conclusions was that "strong links" (i.e links between strongly correlated timeseries) are responsible for clustering while "weak links" provide network connectivity.

Presented in this way, the timeseries clustering task seems very close to graph community detection. Communities in networks are indeed often seen as groups of nodes exhibiting high internal densities of links as well as a low density across communities \cite{Newmanb}. Although several techniques exists (based on different graph properties : modularity \cite{Newmanb}, edge betweeness \cite{Girvan2002}, spectrum of Laplacian \cite{Newman}...), the clique percolation algorithm \cite{Lancichinetti} uses directly this observation and the idea behind it is actually quite simple. It starts indeed by searching for cliques (a complete subgraph) in the given graph as potential community cores. Because other nodes can naturally belong to a community with a less strong connectivity requirement, the algorithm uses a fitness function that quantifies the quality (according to some criterions) of a community structure inside the graph. Expansed cliques alternatively look in their respective neighborhood, select the node that yields the best increase in fitness, and finally incorporate it. The percolation algorithm stops when seeds cease to grow or when all nodes are affected to at least one seed. At this point, some coalitions may be partially overlapping, meaning that a node can belong to multiple communities. Detection of overlapping communities is actually a very active field of research, especially in social networks where persons often belongs to several communities (family, friends, work colleagues...).

@@

\section{Model}
\label{sec:model}

A major concern while designing our model was its ability to translate correctly realistic patterns (consumption and production) as well as realistic geographical correlations between patterns. Largely because of these reasons (but also because they are easily, and sometimes freely, available online) we chose to use meteorological traces as inputs. In this paper we used namely both french data from 2006 to 2012 sampled every 3 hours \cite{Infoclimat} (similar data for the United States covering year 2010 can also be found at \cite{NCDC}). The weather stations already offer a quite good sampling of a given territory, and, with the growth of personal weather stations constantly updating data bases, the mesh becomes finer and finer. As shown in the first part of figure 1, the first step consists in discretizing the studied zone around well chosen weather stations and gathering traces for these stations. In this paper, we consider that prosumers can only produce through wind turbines and photovoltaic panels (PV). On the other hand, we built their consumption patterns according to two major cycles :
\begin{itemize}
\item \textbf{Daily cycles} : Consumption is low during night, and higher during the day with two picks in the morning and evening. Some noise is added so that prosumers have similar but not identical cycles.
\item \textbf{Seasonal cycles} : Consumption is higher in the winter because of heating and low in the summer (we did not consider air conditioning in this paper). Temperature traces were the principal ingredient for modeling these cycles.
\end{itemize}

\begin{center}
\begin{figure}
\includegraphics[scale=0.45]{figure2.jpg}
\caption{Process diagram}
\label{Fig1}
\end{figure}
\end{center}

@@

At this point, modeling agents consists in fixing a few parameters (most of the time drawn from random distributions) such as the geographical position, the number of wind turbines, PV, appliances, the temperature of confort, and so on... The objective is that, depending on the weather of her zone, the DER and appliances she owns, and the way she decides to heat her home, a given prosumer is able to compute her production/consumption at any time. We thus used simple and known power curve models for linking weather variables to output power, such that a wind turbine (or a PV for instance) takes a wind speed (a solar irradiance) as input and gives an instantaneous output power \cite{Kota2011} \cite{windturbinemodel}. Any agent is thus able to compute at any time (within the traces bounds) how much she produces and consumes.

@@

More formally, we denote by $ \mathcal{A} = \{ a_{1},...a_{N} \} $ the set of agents (the prosumers) and $ P_{i}(t) $ the instantaneous power value of agent i at instant t (its instantaneous production minus consumption, meaning that $ P_{i}(t) $ represents agent i available power at instant t). During the simulation (from $t_{0} $ to $ t_{K} $), all agents record their values with an hour time interval. As expected (because we introduced them), the timeseries exhibit high seasonal patterns, but completely different from agent to agent because they depend both on the energetic mix and habits of a given prosumer. Nevertheless, these macro variations can hide the interesting information in the correlation coefficients and twist the rest of our process. Thereby, we remove these seasonal effects for all agents (see second block of figure \ref{Fig1}). We denote by $ \mathcal{T}_{i} = \{ P_{i}(t_{0}),...,P_{i}(t_{K}) \} $ the resulting timeserie for  agent i. For any agent i, we note $ \mathcal{P}_{i} $ the probability distribution drawn from $ \mathcal{T}_{i} $ and by $ P_{i} $ a random variable following $ \mathcal{P}_{i} $.

@@

What this paper investigates in the following is the developement of an algorithm that organizes prosumers such that the synergy term of the standard deviation ( $ \rho_{ij}\sigma_{i}\sigma_{j} $ in the example above) is minimized. In such settings, and through the clique percolation procedure (section \ref{sec:forming}), we will show that coalitions with low relative standard deviation, e.g high utility coalitions, can be computed.

\section{Forming coalitions}
\label{sec:forming}

This section explains the process with which we form the coalitions (see the third block of Figure \ref{Fig1}). First, we need to simulate the timeseries of available power (first two blocks of Figure \ref{Fig1}). We consider a pool $ \mathcal{A} $ of 200 agents, whose parameters were chosen randomly. The prosumers are positioned (also randomly) on a square lattice previously filled with climate vectors obtained from the french data sets (see section \ref{sec:model}). Simulations were run from february 2006 to december 2010 such that we are dealing with 200 hourly sampled timeseries of available power over this period. Removing season trends finally leads us to the formation of coalitions.

The model we used in order to simulate timeseries of available production provide some diversity because of the energetic mix/climate vectors combo. Nevertheless, as the number of agents grows, the timeseries tend to exhibit similar patterns. This is particularly visible when creating a correlation graph $ G_{1}(\mathcal{A},E_{1},\omega_{1}) $ with the same kind of metric as \cite{Garas2008} or \cite{Onnela2004} ($ d_{ij} = 1 - \rho_{ij}^{2} $). Very well defined clusters appear in the corresponding $ \epsilon $-graph for large range of $ \epsilon $.

@@

We thus opt for reversing the metric ($ d_{ij}^{'} = \rho_{ij}^{2} $) such that uncorrelated timeseries are close to each other and (anti)correlated timeseries are distant in $ G_{2}(\mathcal{A},E_{2},\omega_{2}) $. As expected \cite{Onnela2004}, independently of the $ \epsilon $ parameter, the resulting $ \epsilon $-graph exhibits henceforth much less clustering than in $ G_{1} $ and communities seem hardly visible. Therefore, using classical clustering or community detection algorithms do not provide very good results.

@@

As explained in section \ref{sec:related}, clique percolation leads generally to overlapping communities. For simplicity, in this paper, we wish to keep the coalitions separated and  leave the management of overlapping coalitions for future work. We thus implemented a simple heuristic that consider nodes in multiple seeds one by one and chooses its final coalition as the one that “needs it the most” in terms of utility loss. More formally, for a coalition S and a node $ i \in S $, we define :

@@

\section{Results}
\label{sec:results}

\begin{center}
\begin{figure}
\setlength{\tabcolsep}{-0.7em}
\begin{tabular}{cc}
   \includegraphics[scale=0.5]{./figure1/util.eps} &
   \includegraphics[scale=0.5]{./figure6/util2.eps} \\
\end{tabular}
\caption{Utility function in parameters space. Left in $ (\phi, P_{\phi}^{MIN}) $ when $ N_{COAL} $ is fixed. Right in $ (N_{COAL}, P_{\phi}^{MIN}) $ for a fixed $ \phi $}
\label{Fig2}
\end{figure}
\end{center}

As visible on the left plot of figure \ref{Fig2}, the parameters $\phi$ and $P_{\phi}^{MIN}$ shape the utility function such that, if $ \phi $ is close to zero, the reliability requirement is very high and only small values of $ P_{\phi}^{MIN}$ could potentially lead to valid coalitions (and positive utility values). Conversely, the higher $\phi$, the less constraints are imposed to the coalitions and valid ones can arise for a larger spectrum of $ P_{\phi}^{MIN}$. Obviously, the highest utility values are found for high $ \phi $, because coalitions are able then to announce higher contract values, yielding higer utilities.

@@

On the right plot of figure \ref{Fig2}, we can see that when the grid constraints are fixed ($ \phi = 0.08 $ and $ P_{\phi}^{MIN} = 2 $ for instance) and only the number of coalitions varies, there is an increase in social welfare (sum of coalitions utilities) up to a maximum point (for $ N_{COAL} = 3 $ here) before a decrease. The reason is that increasing the number of coalitions allows more coalitions to reach stability and enter the market, but there is a point where nodes bringing stability are not sufficient inside the coalitions to make them pass the grid requirements, and some coalitions start to fail with zero utility. Moreover, reckon that increasing $ N_{COAL} $ means also increasing $ \epsilon $, leading to denser graphs where information is flooded, meaning that the algorithm performances decrease.
Naturally, it is also visible that small values of $P_{\phi}^{MIN}$ lead to higher utilities because coalitions are able to announce higher contract values.

\begin{center}
\begin{figure}[b]
  \includegraphics[scale=0.8]{./figure5/coals.pdf}
  \caption{Evolution of the coalitions for different values of grid requirement $ P_{MIN} $. Blue dots represents clique percolation results while red dots stands for random process results.}
  \label{Fig3}
\end{figure}
\end{center}

For our 200 prosumers, we suppose that the grid fixes a reliability $ \phi = 0.1 $ (meaning that coalitions should produce more than their contract values at least $ 90 \% $ of the time). We also fix the number of coalitions to 10 and have a look at how these coalitions evolve as the grid changes the $ P_{MIN} $ requirement. As a comparison, we use a completely random algorithm that only asks for a number of coalitions and partitions the agents in a random fashion. Figure \ref{Fig3} shows this evolution for our algorithm (blue dots) and for the random process (red dots). The diameter of a dot is proportional to the number of agents in the coalition and the higher the dot is, the higher its utility. The $ P_{MIN} $ values of the x axis are expressed in tenth of MW for readability and the hatched zone corresponds to the "under-requirement space", meaning that whenever a coalition is in this zone, it has a null utility.

Looking at figure \ref{Fig3}, we can see first that our algorithm seems to make generally a better job at finding high valued coalitions, but also that the results seem far more robust against how the grid positions its requirements. The blue dots allowed to enter the market seem indeed to outnumber the red dots, especially when the grid requirements are neither to low or to high ($ P_{MIN} = 5 $ in figure \ref{Fig3}).

In order to explore a little bit more in this direction, we have a look, on figure \ref{Fig4}, at the percentage of coalitions able to enter the market for different values of $ P_{MIN} $. For consistency, we always average the results of the random procedure over 100 realizations (5 realizations for our algorithm, but the results are relatively stable), and the errorbars in the plots stands for the standard deviations of the results.

@@

\begin{center}
\begin{figure}
   \includegraphics[scale=0.7]{./figure4/survival.eps}
   \caption{Percentage of coalitions able to enter the market for different values of grid requirement $ P_{MIN} $. Blue curve represents clique percolation results while red curve stands for random process results}
   \label{Fig4}
\end{figure}
\end{center}

@@

\section{Conclusion}
\label{sec: conclusion}

In this paper we collected meteorological traces for weather stations sampling a given territory. After discetization of the space, each zone receives the climate vector corresponding to its mother weather station. Prosumers are then distributed randomly between zones and each prosumer receives a random number of DER and loads as well as diverse consumption habits. Simulations are then runned and allow us to obtain, for each prosumer, the timeserie describing the instantaneous power that he could potentially be wanted to sell on the market. Nevertheless, as entering on the energy market is constrained by the grid operator, such that only sufficiently reliable and sufficiently productive units are allowed to enter, prosumers have usually no other choice than forming coalitions. We then introduced a framework such that the grid requirements take a precise meaning and such that coalitions announce contract values accordingly.

Through this paper, our interest was on how we could circumscribe groups of prosumers that can rise above the grid requirements. In this direction, we removed seasonal variations from the prosumers timeseries and proposed to organize them over a correlation graph with a reversed metric. A greedy clique percolation process on the $ \epsilon $-graph enables us to finally establish coalitions of agents. We showed in section \ref{sec:results} that our algorithm was actually finding coalitions with high utility values but also that it was far more robust than a random search over coalition structures when the grid operator increases his requirements (Figure \ref{Fig4}).
